---
title: 'The seeds of reproducibility - a 2 minute crash course'
author: 'Matt Cole'
date: '2017-08-25'
slug: 'seeds'
categories: ['R']
tags: ["reproducibility", "seed"]
---

While a Teaching Assistant for the 620 [Biostatistics sequence](http://www.jhsph.edu/courses/course/17963/2013/140.623.01/statistical-methods-in-public-health-iii) at JHU (Statistical Methods in Public Health) one issue came up fairly regularly, the idea of seeds and reproducing results when sometimes non-obvious random elements are involved.

Many analyses are dependent on some sort of random number generation, weather it be from cross-validation, bootstrapping, imputing, or simulating, it's inevitable that you'll need these processes. Because of the nature of 'random' things, reproducibility can be tough. How can you reproduce a dice roll of 5? 

R doesn't utilize true random numbers, but rather pseudo random numbers, which are sequences of numbers with properties of truly random number but via a deterministic process. That is, the pseudo random numbers are already determined before we access them. Which 'state' of pseudo random number generation we utilize is determined entirely by this seed.

So when we call a random normal, we cannot reproduce that result easily:

```{r echo = F}
set.seed(20170707)
```

#### Simulation 1

```{r}
rnorm(1)
```

Now, if we were to re run the simulation, we would get different results:

#### Simulation 2

```{r}
rnorm(1)
```

Instead, by calling a seed right before, we can guarantee the same results.
The way seeds are implemented in statistical packages are a bit different, like so:

 * Python: `import random` and `random.seed(n)`
 * R: `set.seed(n)`
 * STATA: `set seed n`

where n is some number (integer) of our choice.

#### Simulation 1A

```{r}
set.seed(100)
rnorm(1)
```

#### Simulation 2A

```{r}
set.seed(100)
rnorm(1)
```

Notice here, we get the same exact result - perfect for reproducibility and consistency. 

There is a ton of information out there on this subject (this was just the crash course I give students). I will attach some links below.

Enjoy!

### Links:
 * [Pseudorandom number generator](https://en.wikipedia.org/wiki/Pseudorandom_number_generator) (wiki link)
 * [Pseudorandomness](https://en.wikipedia.org/wiki/Pseudorandomness) (wiki link)
 * [set.seed help](https://stat.ethz.ch/R-manual/R-devel/library/base/html/Random.html) (R help)
 * [Random v. Pseudorandom](https://boallen.com/random-numbers.html) (Blog)


